---
title: "CPS - IPMUS: Paid Childcare Workers"
author: "Katherine Rodriguez"
format: 
  html:
    toc: true
    toc-location: left
    number-sections: true
    theme: cosmo
editor: visual
---

## Table of Contents

0.  Introduction/background research
1.  Setup
2.  Load and prepare data
3.  Build survey designs
4.  Descriptive results (paid care)
5.  Dollar valuation
6.  Paid childcare (BLS/CPS)
7.  Plots / tables
8.  Concluding thoughts

## Introduction

Childcare is an undervalued service in both unpaid (household labor) and paid work (childcare workforce). Both are closely connected, families supply time, while the childcare industry supplies services. The objective of this project is to estimate the dollar value of unpaid childcare and measure it against market pay for childcare workers, demonstrating that childcare is undervalued both within households and the childcare industry.

**Research Question 2 (comparative)** Among paid childcare workers, how do wages compare to similar occupations (Preschool teachers, Home and Health Aides)? What is the gender composition and race/ethnicity profile of unpaid and paid childcare work?

## Setup

Load necessary libraries.

```{r, message=FALSE, warning=FALSE}
# Packages, survey options, and BLS wage constants live in R/setup.R.
source(here::here("R", "setup.R"))
```

**Survey Defaults**

Survey variance options (`survey.lonely.psu`, `scipen`) are set once in
`R/setup.R`, sourced above.

## Load and Prepare Data

The data is loaded into the program to prepare for analyses.

```{r}
# R automatically decompresses .gz files
# CPS raw extract downloaded from IPUMS-CPS. Place the file at
# data/raw/cps_ipums.csv.gz (see data/raw/README.md for how to recreate
# the extract).
cps_raw <- read_csv(here::here("data", "raw", "cps_ipums.csv.gz"))
dim(cps_raw)
```

**Keep ASEC-only observations/Clean Variables**

```{r}
cps <- cps_raw |>
  filter(
    YEAR == 2023,
    ASECFLAG == 1
    ) |>
  
  #Key variables as numeric
  mutate(
    AGE       = as.numeric(AGE),
    SEX       = as.numeric(SEX),
    RACE      = as.numeric(RACE),
    HISPAN    = as.numeric(HISPAN),
    EMPSTAT   = as.numeric(EMPSTAT),
    OCC       = as.numeric(OCC),
    ASECWT    = as.numeric(ASECWT),
    INCWAGE   = as.numeric(INCWAGE),
    UHRSWORKT = as.numeric(UHRSWORKT),

    #Convert replicate weights to numeric
    across(matches("^REPWTP[0-9]+$"), as.numeric),
    
    #Clean INCWAGE Variable
    INCWAGE = ifelse(INCWAGE %in% c(99999999, 99999998), NA_real_, INCWAGE)
  ) |>
    
    #Trim dataset to relevant variables
    select(
      YEAR, ASECWT, ASECFLAG, matches("^REPWTP[0-9]+$"),
      AGE, SEX, RACE, HISPAN, EDUC,
      EMPSTAT, OCC, INCWAGE, UHRSWORKT,
  )

dim(cps)
summary(cps$INCWAGE)
```

**Save an .rds Version for Faster Reloads**

```{r}
saveRDS(cps, here::here("data", "processed", "cps_2023_childcare.rds"))
```

```{r}
cps <- readRDS(here::here("data", "processed", "cps_2023_childcare.rds"))
head(cps)
```

**Recode Sex, Race/Ethnicity and Occupations**

```{r}
#Clean race up a bit
cps <- cps |>
  mutate(
    RACE = ifelse(RACE == 999, NA_integer_, RACE)
  )


cps <- cps |>
  mutate(
    #Recode sex category
    sex_category = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE     ~ NA_character_
    ),
    female = (sex_category == "Female"),
    male   = (sex_category == "Male"),
    
    #Recode race/ethnicity
    hispanic_any = case_when(
      HISPAN == 0 | HISPAN == 100 ~ 0L,
      HISPAN > 0 & HISPAN < 900   ~ 1L,
      TRUE                        ~ NA_integer_
    ),
    white_nh  = ifelse(hispanic_any == 0L & RACE == 100, 1L, 0L),
    black_nh  = ifelse(hispanic_any == 0L & RACE == 200, 1L, 0L),
    native_nh = ifelse(hispanic_any == 0L & RACE == 300, 1L, 0L),
    asian_nh  = ifelse(
    hispanic_any == 0L & RACE %in% c(650, 651, 652), 1L, 0L 
    ),
    other_nh = ifelse(
      hispanic_any == 0L &
        RACE %in% c(
          700,     #other single rae, n.e.c.
          801:820, #2-3 race combos (with various combos)
          830      #4-5 race combos
        ),
      1L, 0L
    ),
    
    #Collapsed race/ethnicity category
    race_cat = case_when(
      hispanic_any == 1L          ~ "Hispanic",
      white_nh     == 1L          ~ "White",
      black_nh     == 1L          ~ "Black",
      asian_nh     == 1L          ~ "Asian",
      native_nh    == 1L          ~ "Native American",
      other_nh     == 1L          ~ "Other/Multiracial",
      TRUE                         ~ NA_character_
    ),
    
    #Recode occupation groups
    childcare_worker  = (OCC == 4600),
    pre_kinder_teach  = (OCC == 2300),
    #home_health_aide  = (OCC %in% c(3601,3602)), #not needed for this project, but could be added back in if we want to expand to other care occupations
    
    #Collapsed occupation category
    occ_group = case_when(
      childcare_worker  ~ "Childcare Workers",
      pre_kinder_teach  ~ "Preschool/ K Teachers",
      #home_health_aide  ~ "Home Health & Personal Care Aides", removing home health aides for this project to focus on childcare occupations, but could be added back in if we want to expand to other care occupations
      TRUE              ~ "Other"
    ),
    
    #Factor sex category
    sex_category = factor(
     sex_category, levels = c("Female", "Male")  
    ),
    
    #Factor race_cat
    race_cat = factor(
     race_cat,
     levels = c(
       "White",
       "Black",
       "Hispanic",
       "Asian",
       "Native American",
       "Other/Multiracial"
    )
  ),
    #Factor occupation group
    occ_group = factor(
      occ_group,
      levels = c(
        "Childcare Workers",
        "Preschool/ K Teachers",
        #"Home Health & Personal Care Aides", removing for now
        "Other"
    )
  )
)

#Check recode
table(cps$sex_category, useNA = "ifany")

```

**Filter to Workers Aged 16 and Older with Positive Wages**

```{r}
#Define emplyed workers with positive wages
cps_workers <- cps |>
  mutate(
    employed    = EMPSTAT %in% c(10,12),
    wage_annual = ifelse(INCWAGE > 0, INCWAGE, NA_real_)
  ) |>
  
  #Filter to age > 16
  filter(
    AGE >= 16,
    employed,
    !is.na(wage_annual)
  )

cps_workers <- cps_workers |>
  mutate(
    full_time = UHRSWORKT >= 35
  )
cps_workers
```

## Build Survey Designs

```{r}
design_cps <- svrepdesign(
  data             = cps_workers,
  weights          = ~ASECWT,
  repweights       = "REPWTP[0-9]+$",
  type             = "successive-difference",
  combined.weights = TRUE,
  mse              = TRUE
)

design_cps <- update(
  design_cps,
  full_time = cps_workers$full_time
)

# Mean wages by occupation & race, full-time only
svyby(
  ~wage_annual,
  ~occ_group + race_cat,
  subset(design_cps, full_time),
  svymean,
  na.rm = TRUE
)
```

## Descriptive Results

**Composition of Sex Category in Each Occupation**

```{r}
sex_comp <- svyby(
  ~sex_category,
  ~occ_group, 
  design_cps,
  svymean,
  na.rm = TRUE
)
sex_comp
```

**Race/Ethnicity Composition by Occupation**

```{r}
race_comp <- svyby(
  ~race_cat,
  ~occ_group, 
  design_cps,
  svymean,
  na.rm = TRUE
)
race_comp

#Proportion of full_time workers by occupation and race
svyby(
  ~full_time,
  ~occ_group + ~race_cat,
  design_cps,
  svymean,
  na.rm = TRUE
)
```

**Mean Wages by Occupation**

```{r}
mean_wage_occ <- svyby(
  ~wage_annual,
  ~occ_group,
  design_cps,
  svymean,
  na.rm = TRUE
)
mean_wage_occ
```

**Mean Wage by Occupation and Sex**

```{r}
mean_wage_sex <- svyby(
  ~wage_annual,
  ~occ_group + sex_category,
  design_cps,
  svymean,
  na.rm = TRUE
)
mean_wage_sex
```

**Occupation Mean Wages by Race/Ethnicity**

```{r}
mean_wage_r <- svyby(
  ~wage_annual,
  ~occ_group + race_cat,
  design_cps,
  svymean,
  na.rm = TRUE
)
mean_wage_r
```

## Regression

Preliminary

```{r}

cps_reg <- svyglm(
  wage_annual ~ occ_group + sex_category + race_cat + full_time + AGE,
  design = design_cps
)

summary(cps_reg)

```

**Add Sex x Occupation Interaction**

```{r}
cps_reg_int <- svyglm(
  wage_annual ~ occ_group * sex_category + race_cat + full_time + AGE,
  design = design_cps
) 

summary(cps_reg_int)

# Check the levels used in the model/design
levels(design_cps$variables$occ_group)
levels(design_cps$variables$sex_category)
levels(design_cps$variables$race_cat)
class(design_cps$variables$full_time)

```

**Predicted Wages for Each Group**

```{r, eval=FALSE}
newdat <- expand.grid(
  occ_group    = c("Childcare Workers", "Preschool/ K Teachers", "Other"),
  sex_category = c("Female", "Male"),
  race_cat     = "White",          # fix race to simplify story
  full_time    = TRUE,             # full-time workers
  AGE          = 30                # pick a representative age
)

preds <- predict(
  cps_reg_int,
  newdata = newdat,
  se.fit  = TRUE
)

pred_df <- newdat |>
  mutate(
    wage_pred = as.numeric(preds$fit),
    wage_se   = as.numeric(preds$se.fit),
    wage_lo   = wage_pred - 1.96 * wage_se,
    wage_hi   = wage_pred + 1.96 * wage_se
  )

pred_df

#Plot
ggplot(pred_df, aes(x = occ_group, y = wage_pred,
                    fill = sex_category)) +
  geom_col(position = position_dodge(width = 0.6)) +
  geom_errorbar(aes(ymin = wage_lo, ymax = wage_hi),
                position = position_dodge(width = 0.6),
                width = 0.2) +
  labs(
    x = "Occupation group",
    y = "Predicted annual wage (US$)",
    fill = "Sex",
    title = "Predicted annual wages by occupation and sex\n(White, age 35, full-time, CPS 2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format())


```

## Summary Table

**Key Statistics by Occupation**

```{r}
## 1. Mean wage by occupation ----
wage_occ <- svyby(
  ~ wage_annual,
  ~ occ_group,
  design_cps,
  svymean,
  na.rm = TRUE
) |>
  as.data.frame() |>
  rename(
    mean_wage = wage_annual,
    se_wage   = se
  )

## 2. Sex composition by occupation ----
# (Using the structure you showed earlier:
#  columns I(sex_category)Female, I(sex_category)Male, se1, se2)

sex_occ <- svyby(
  ~ sex_category,
  ~ occ_group,
  design_cps,
  svymean,
  na.rm = TRUE
) |>
  as.data.frame() |>
  mutate(
    pct_female = sex_categoryFemale * 100,
    pct_male   = sex_categoryMale   * 100
  ) |>
  select(occ_group, pct_female, pct_male)

## 3. Full-time share by occupation ----
ft_occ <- svyby(
  ~ full_time,
  ~ occ_group,
  design_cps,
  svymean,
  na.rm = TRUE
) |>
  as.data.frame() |>
  mutate(
    pct_fulltime = full_timeTRUE * 100
  ) |>
  select(occ_group, pct_fulltime)

## 4. Join everything into one CPS summary table ----
cps_table <- wage_occ |>
  select(occ_group, mean_wage, se_wage) |>
  left_join(sex_occ, by = "occ_group") |>
  left_join(ft_occ, by = "occ_group")

cps_table
```

**Put Table into Kable Format**

```{r}
cps_table_display <- cps_table |>
  mutate(
    # pretty formatting
    mean_wage_fmt = sprintf("$%s", format(round(mean_wage, 0), big.mark = ",")),
    se_wage_fmt   = sprintf("$%s", format(round(se_wage,   0), big.mark = ",")),
    pct_female_fmt   = sprintf("%.1f%%", pct_female),
    pct_male_fmt     = sprintf("%.1f%%", pct_male),
    pct_fulltime_fmt = sprintf("%.1f%%", pct_fulltime)
  ) |>
  select(
    `Occupation group` = occ_group,
    `Mean annual wage` = mean_wage_fmt,
    `SE (wage)`        = se_wage_fmt,
    `% female`         = pct_female_fmt,
    `% male`           = pct_male_fmt,
    `% full-time`      = pct_fulltime_fmt
  )

kable(
  cps_table_display,
  caption = "Summary of paid care and other occupations, CPS ASEC 2023",
  align   = c("l", "r", "r", "r", "r", "r")
) |>
  kable_styling(full_width = FALSE)
```

## Visualizations

**Mean Wages by Occupation Plot**

```{r}
mean_wage_occ <- svyby(
  ~wage_annual,
  ~occ_group,
  design_cps,
  svymean,
  na.rm = TRUE
)|> 
 data.frame()|>

 mutate(
    mean_wage = wage_annual,
    se_wage   = se
  )
mean_wage_occ
ggplot(mean_wage_occ, aes(x = occ_group, y = mean_wage)) +
  geom_col(fill = "#6C4B9E") +
  #coord_flip() +
  geom_errorbar(aes(ymin = mean_wage - 1.96 * se_wage, 
                    ymax = mean_wage + 1.96 * se_wage),
                width = 0.2) +
  labs(
    x = "Occupation Group",
    y = "Mean Annual Wage (USD)",
    title = "Mean Annual Wages by Occupation Group",
    ) + 
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 20, hjust = 1), 
   plot.title   =  element_text(hjust = 0.5),    
   panel.border = element_rect(color = "black", fill = NA, linewidth =1))
```

**Sex Composition by Occupation Plot**

```{r}
sex_comp_long <- sex_comp |>
  tibble::as_tibble() |>
  tidyr::pivot_longer(
    cols = c(sex_categoryFemale, sex_categoryMale),
    names_to = "sex",
    values_to = "share"
  ) |>
  mutate(
    sex = ifelse(grepl("Female", sex), "Female", "Male"),
    share_pct = share * 100
  )

ggplot(sex_comp_long, aes(x = occ_group, y = share_pct, fill = sex)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Female" = "#2C3E50", "Male" = "#1ABC9C")) +
  geom_text(
    aes(label = paste0(round(share_pct, 1), "%")),
    position = position_dodge(width = 0.7),
    hjust = -0.2,
    fontface = "bold",
    color = "grey30",
    size = 3.2
  ) +
  labs(
    x = NULL,
    y = "Share of workers",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom"
  )
```

**Race/Ethnicity Composition by Occupation**

```{r}
race_comp_long <- race_comp |>
  mutate(
    pct_white     =  race_catWhite    * 100,
    pct_black     =  race_catBlack    * 100,
    pct_hispanic  =  race_catHispanic * 100,
    pct_asian     =  race_catAsian    * 100,
    pct_native    = `race_catNative American`   * 100,
    pct_other     = `race_catOther/Multiracial` * 100
  ) |>
  select(occ_group, starts_with("pct_")) |>
  pivot_longer(
    cols      = starts_with("pct_"),
    names_to  = "race",
    values_to = "pct"
  )

ggplot(race_comp_long, aes(x = occ_group, y = pct, fill = race))+
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    x = "Occupation Group",
    y = "Composition",
    title = "Racial/Ethnicity Composition of Childcare and Care Occupation"
  )+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
   plot.title   =  element_text(hjust = 0.5),    
   panel.border = element_rect(color = "black", fill = NA, linewidth =1)) 

```

**Mean Wages by Occupation and Sex**

```{r}
mean_wage_sex <- svyby(
  ~wage_annual,
  ~occ_group + sex_category,
  design_cps,
  svymean,
  na.rm = TRUE
)|>
  mutate(
    mean_wage = wage_annual,
    se_wage   = se
  )
mean_wage_sex
ggplot(mean_wage_sex, aes(x = occ_group, y = mean_wage, fill = sex_category)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = mean_wage - 1.96 * se_wage, 
                    ymax = mean_wage + 1.96 * se_wage),
                position = position_dodge(width = 0.9),
                width = 0.2) +
  scale_fill_manual(values = c("Female" = "#6C4B9E", "Male" = "grey40")) +
  labs(
    x = "Occupation Group",
    y = "Mean Annual Wage (USD)",
    title = "Gender Wage Differences by Occuptation"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
  plot.title   =  element_text(hjust = 0.5),    
  panel.border = element_rect(color = "black", fill = NA, linewidth =1))
```

## Dollar Valuation of Unpaid Childcare

**Implied Annual Value of Unpaid Care**

```{r}
#Mean childcare rate (defined once in R/setup.R)
childcare_rate <- BLS_WAGE_CHILDCARE_WORKER

#Average hours per day of childcare for primary caregivers.
#This value is produced by 01_unpaid_childcare_atus.qmd, which writes it to
#data/processed/atus_primary_hours.rds. Run that notebook first. Reading the
#saved artifact (instead of copying the number by hand) keeps the two
#notebooks in sync.
atus_primary_hours <- readRDS(
  here::here("data", "processed", "atus_primary_hours.rds")
)

mean_cc_hrs_post <- atus_primary_hours$mean_cc_hrs
se_cc_hrs_post   <- atus_primary_hours$se_cc_hrs
as.numeric(mean_cc_hrs_post)

unique(mean_wage_occ$occ_group)

#Implied annual value of unpaid childcare
unpaid_val_annual <- mean_cc_hrs_post * childcare_rate * 365
unpaid_val_annual

childcare_annual  <- mean_wage_occ$mean_wage[mean_wage_occ$occ_group == "Childcare Workers"]
prek_annual       <- mean_wage_occ$mean_wage[mean_wage_occ$occ_group == "Preschool/ K Teachers"]
homehealth_annual <- mean_wage_occ$mean_wage[mean_wage_occ$occ_group == "Other"] 


unique(mean_wage_occ$occ_group)
valuation_df <- tibble(
  group = c(
    "Unpaid Primary Caregiver (implied)",
    "Childcare Workers (paid)",
    "Preschool/ K Teachers (paid)",
    "Other (paid)"
  ),

  annual_value = c(
    unpaid_val_annual,
    childcare_annual,
    prek_annual,
    homehealth_annual
  )
)

valuation_df <- valuation_df |>
  mutate(
    annual_value_k = annual_value / 1000
  )

valuation_df

# Color palette: unpaid highlighted, paid uniform
val_colors <- c(
  "Unpaid Primary Caregiver (implied)" = "#6C4B9E",   # highlight purple
  "Childcare Workers (paid)"           = "#8FB3D9",   # uniform blue
  "Preschool/ K Teachers (paid)"       = "#8FB3D9",   # uniform blue
  "Other (paid)"                       = "#8FB3D9"    # uniform blue
) 


#Bar Plot
ggplot(valuation_df, aes(x = group, y = annual_value_k))+ 
  geom_col(col = "grey40") +
  geom_text(
    aes(label = paste0("$", round(annual_value_k, 1), "k")),
    position = position_dodge(width = 0.7),
    hjust = 1.2,
    fontface = "bold",
    color = "white",
    size = 3.2
  ) +
  scale_fill_manual(values = val_colors) +
  labs(
    x = NULL,
    y = "Annual Value (in thousands USD)"
    #subtitle = "Unpaid Value Based on Atus Hours (2021-2024) \n x Childcare Worker Hourly Wage"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
  plot.title   =  element_text(hjust = 0.5))
```

```{r}
mean_wage_occ$occ_group
```

**Implied Annual Value of Unpaid Care B**

```{r}

# Color palette: unpaid highlighted, paid uniform
val_colors <- c(
  "Unpaid Primary Caregiver (implied)" = "#1ABC9C",   # teal
  "Childcare Workers (paid)"           = "#2C3E50",   # uniform blue
  "Preschool/ K Teachers (paid)"       = "#2C3E50",   # uniform blue
  "Other (paid)"                       = "#2C3E50"    # uniform blue
)

# Ensure ordering
valuation_df$group <- factor(
  valuation_df$group,
  levels = c(
    "Unpaid Primary Caregiver (implied)",
    "Childcare Workers (paid)",
    "Preschool/ K Teachers (paid)",
    "Other (paid)"
  )
)

# Plot
ggplot(valuation_df, aes(x = group, y = annual_value_k, fill = group)) +
  geom_col(color = "grey40") +
  geom_text(
    aes(label = paste0("$", round(annual_value_k, 1), "k")),
    hjust = 1.2,
    fontface = "bold",
    color = "white",
    size = 3.2
  ) +
  scale_fill_manual(values = val_colors) +
  labs(
    x = NULL,
    y = "Annual Value (in thousands USD)"
  ) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(
    plot.title   = element_text(hjust = 0.5),
    legend.position = "none"
  )


```
